---
title: 'Ridge Regression'
output:
  html_document:
    code_folding: 'hide'
---

<br>

**Change:** Model is run multiple times with different train and test sets

```{r load_packages, warning=FALSE, message=FALSE}

library(tidyverse)
library(knitr)
library(plotly) ; library(viridis) ; library(gridExtra) ; library(RColorBrewer)
library(biomaRt)
library(Rtsne)
library(caret) ; library(ROCR) ; library(car)
library(corrplot)
library(expss) ; library(knitr)

SFARI_colour_hue = function(r) {
  pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3','#808080','gray','#d9d9d9')[r]
}
```

## Background
<br>

In 20_04_07_LR.html we performed a logistic regression but we discovered that the features were strongly correlated. This inflates the standard error of the coefficients, making them no longer interpretable.

```{r, warning=FALSE, message=FALSE}

# Clusterings
clustering_selected = 'DynamicHybridMergedSmall'
clusterings = read_csv('./../Data/clusters.csv')
clusterings$Module = clusterings[,clustering_selected] %>% data.frame %>% unlist %>% unname

assigned_module = data.frame('ID' = clusterings$ID, 'Module' = clusterings$Module)

# Original dataset
original_dataset = read.csv(paste0('./../Data/dataset_', clustering_selected, '.csv'), row.names=1)

# Model dataset
load('./../Data/LR_model.RData')

# Add gene symbol
getinfo = c('ensembl_gene_id','external_gene_id')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl',
               host='feb2014.archive.ensembl.org')
gene_names = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=rownames(dataset), mart=mart)
```

Variance Inflation Factor (VIF) and correlation plot
```{r, fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
# VIF
plot_data = data.frame('Feature' = car::vif(fit) %>% sort %>% names,
                       'VIF' = car::vif(fit) %>% sort %>% unname) %>%
            mutate(outlier = VIF>10)

plot_data %>% ggplot(aes(reorder(Feature, -VIF), VIF, fill = !outlier)) + geom_bar(stat='identity') + scale_y_log10() +
              geom_hline(yintercept = 10, color = 'gray', linetype = 'dashed') + xlab('Model Features') + theme_minimal() +
              theme(legend.position = 'none', axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
# Correlation plot
corrplot.mixed(cor(train_set[,-ncol(train_set)]), lower = 'number', lower.col = 'gray', number.cex = .6, tl.pos = 'l', tl.col = '#666666')


rm(getinfo, mart, clusterings)
```

### Possible solutions to Multicollinearity:
<br>

1. Remove all variables with a VIF>10: We would lose all but three of our variables, not ideal

2. Do Principal Component Regression: We would lose the relation between the prediction and the original features, which would be interesting to study

3. Don't do anything: Multicollinearity affects the coefficients and p-values of the regression, but it doesn't affect the predictions, precision of the predictions or the goodness-of-fit statistics [ref](https://statisticsbyjim.com/regression/multicollinearity-in-regression-analysis/), but as with the previous option, we cannot study the coefficients of the regression

4. **Use Ridge Regression**

<br>

## Ridge Regression

<br>

Notes:

- Running the model multiple times to get more acurate measurements of its performance

- Over-sampling positive samples in the training set to obtain a 1:1 class ratio

```{r train_model}

### DEFINE FUNCTIONS

# Create positive training set including all SFARI scores
positive_sample_balancing_SFARI_scores = function(p, seed){
  
  set.seed(seed)
  positive_train_idx = c()
  
  for(score in 1:6){
    score_genes = rownames(original_dataset)[rownames(original_dataset) %in% rownames(dataset) & original_dataset$gene.score == score]
    score_idx = which(rownames(dataset) %in% score_genes)
    score_train_idx = sample(score_idx, size = ceiling(p*length(score_idx)))
    
    positive_train_idx = c(positive_train_idx, score_train_idx)
  }
  
  return(positive_train_idx)
}

create_train_test_sets = function(p, over_sampling_fold, seed){
  
  ### CREATE POSITIVE TRAINING SET (balancing SFARI scores and over-sampling)
  positive_train_idx = positive_sample_balancing_SFARI_scores(p, seed)
  add_obs = sample(positive_train_idx, size = ceiling(over_sampling_fold*length(positive_train_idx)), replace=TRUE)
  positive_train_idx = c(positive_train_idx, add_obs)
  
  
  ### CREATE NEGATIVE TRAINING SET
  negative_idx = which(!dataset$SFARI)
  negative_train_idx = sample(negative_idx, size = length(positive_train_idx))
  
  
  ### CREATE TRAIN AND TEST SET
  train_set = dataset[sort(c(positive_train_idx, negative_train_idx)),]
  test_set = dataset[-unique(c(positive_train_idx, negative_train_idx)),]
  
  return(list('train_set' = train_set, 'test_set' = test_set))
  
}

run_model = function(p, over_sampling_fold, seed){
  
  # Create train and test sets
  train_test_sets = create_train_test_sets(p, over_sampling_fold, seed)
  train_set = train_test_sets[['train_set']]
  test_set = train_test_sets[['test_set']]
  
  # Train Model
  train_set$SFARI = train_set$SFARI %>% as.factor
  lambda_seq = 10^seq(1, -4, by = -.1)
  set.seed(123)
  fit = train(SFARI ~., data = train_set, method = 'glmnet', trControl = trainControl('cv', number = 10),
                tuneGrid = expand.grid(alpha = 0, lambda = lambda_seq))
  
  # Predict labels in test set
  predictions = fit %>% predict(test_set, type='prob')
  preds = data.frame('ID'=rownames(test_set), 'prob'=predictions$`TRUE`) %>% mutate(pred = prob>0.5)

  # Measure performance of the model
  acc = mean(test_set$SFARI==preds$pred)
  pred_ROCR = prediction(preds$prob, test_set$SFARI)
  AUC = performance(pred_ROCR, measure='auc')@y.values[[1]]
  
  # Extract coefficients from features
  coefs = coef(fit$finalModel, fit$bestTune$lambda) %>% as.vector
  
  return(list('acc' = acc, 'AUC' = AUC, 'preds' = preds, 'coefs' = coefs))
}


### RUN MODEL

# Parameters
p = 0.8
over_sampling_fold = 3
n_iter = 100
seeds = 123:(123+n_iter-1)

# Store outputs
acc = c()
AUC = c()
predictions = data.frame('ID' = rownames(dataset), 'SFARI' = dataset$SFARI, 'prob' = 0, 'pred' = 0, 'n' = 0)
coefs = data.frame('var' = c('Intercept', colnames(dataset[,-ncol(dataset)])), 'coef' = 0)

for(seed in seeds){
  
  # Run model
  model_output = run_model(p, over_sampling_fold, seed)
  
  # Update outputs
  acc = c(acc, model_output[['acc']])
  AUC = c(AUC, model_output[['AUC']])
  preds = model_output[['preds']]
  coefs$coef = coefs$coef + model_output[['coefs']]
  update_preds = preds %>% dplyr::select(-ID) %>% mutate(n=1)
  predictions[predictions$ID %in% preds$ID, c('prob','pred','n')] = predictions[predictions$ID %in% preds$ID, c('prob','pred','n')] + update_preds
}

coefs = coefs %>% mutate(coef = coef/n_iter)
predictions = predictions %>% mutate(prob = prob/n, pred_count = pred, pred = prob>0.5)

rm(p, over_sampling_fold, seeds, update_preds, positive_sample_balancing_SFARI_scores, create_train_test_sets, run_model)

```

<br>

To summarise in a single value the predictions of the models:

- prob = Average of all the probabilities

- pred = 1 if prob>0.5, 0 otherwise

```{r, warning=FALSE, message=FALSE}

test_set = predictions %>% filter(n>0) %>% left_join(dataset %>% mutate(ID = rownames(.)) %>% dplyr::select(ID, GS, MTcor), by = 'ID')
rownames(test_set) = predictions$ID[predictions$n>0]

```

<br>

### Performance metrics

<br>

#### Confusion matrix
```{r}
conf_mat = test_set %>% apply_labels(SFARI = 'Actual Labels', 
                                     prob = 'Assigned Probability', 
                                     pred = 'Label Prediction')

cro(conf_mat$SFARI, list(conf_mat$pred, total()))

rm(conf_mat)
```

#### Accuracy

```{r}
cat(paste0('ACCURACY: mean = ', round(mean(acc),4), ' SD = ', round(sd(acc),4)))
```
<br>

#### ROC Curve
```{r ROC_curve}
cat(paste0('AUC:      mean = ', mean(AUC), ' SD = ', sd(AUC)))

pred_ROCR = prediction(test_set$prob, test_set$SFARI)

roc_ROCR = performance(pred_ROCR, measure='tpr', x.measure='fpr')
auc = performance(pred_ROCR, measure='auc')@y.values[[1]]

plot(roc_ROCR, main=paste0('ROC curve (AUC=',round(auc,2),')'), col='#009999')
abline(a=0, b=1, col='#666666')
```

#### Lift Curve
```{r lift_plot}
lift_ROCR = performance(pred_ROCR, measure='lift', x.measure='rpp')
plot(lift_ROCR, main='Lift curve', col='#86b300')

rm(pred_ROCR, roc_ROCR, AUC, lift_ROCR)
```

<br>

---

<br>

### Coefficients

<br>

MTcor has a very small coefficient and Gene Significance has a negative coefficient (!)

```{r, warning=FALSE, message=FALSE}
gene_corr_info = dataset %>% mutate('ID' = rownames(dataset)) %>% dplyr::select(ID, MTcor, SFARI) %>% left_join(assigned_module, by ='ID') %>%
                 mutate(Module = gsub('#','',Module))

coef_info = coefs %>% mutate('feature' = gsub('MM.','',var)) %>% left_join(gene_corr_info, by = c('feature' = 'Module')) %>% 
            dplyr::select(feature, coef, MTcor, SFARI) %>% group_by(feature, coef, MTcor) %>% summarise('SFARI_perc' = mean(SFARI)) %>%
            arrange(desc(coef))

kable(coef_info %>% dplyr::select(feature, coef) %>% rename('Feature' = feature, 'Coefficient' = coef),
      align = 'cc', caption = 'Regression Coefficients')
```
<br>

- There is a positive relation between the coefficient assigned to the membership of each module and the percentage of SFARI genes that are assigned to that module

```{r, warning=FALSE, message=FALSE}
ggplotly(coef_info %>% dplyr::rename('Module' = feature) %>% filter(!is.na(MTcor)) %>%
              ggplot(aes(coef, SFARI_perc)) +  geom_smooth(method = 'lm', color = 'gray', alpha = 0.1) + 
              geom_point(aes(id = Module), color = paste0('#',coef_info$feature[!is.na(coef_info$MTcor)])) + 
              theme_minimal() + xlab('Coefficient') + 
              ylab('% of SFARI genes in Module'))
```

<br>

- There doesn't seem to be a relation between the coefficient and the correlation of the module and the diagnosis.

This is not a surprise since we knew that there was a negative relation between SFARI genes and Module-Diagnosis correlation from Preprocessing/Gandal/AllRegions/RMarkdowns/20_04_03_WGCNA_modules_EA.html. The fact that there is no relation between coefficient and Module-Diagnosis correlation could even be a good sign that the model is picking some biological signal as well as the SFARI patterns (since the relation with the biological signals is positive)

```{r, warning=FALSE, message=FALSE}
ggplotly(coef_info %>% dplyr::rename('Module' = feature) %>% filter(!is.na(MTcor)) %>%
              ggplot(aes(coef, MTcor)) +  geom_smooth(method = 'lm', color = 'gray', alpha = 0.1) + 
              geom_point(aes(id = Module), color = paste0('#',coef_info$feature[!is.na(coef_info$MTcor)])) + 
              theme_minimal() + xlab('Coefficient') + 
              ylab('Module-Diagnosis correlation'))
```

<br>

---

<br>

### Analyse model

<br>


SFARI genes have a slightly higher score distribution than the rest
```{r}
plot_data = test_set %>% dplyr::select(prob, SFARI)

ggplotly(plot_data %>% ggplot(aes(prob, fill=SFARI, color=SFARI)) + geom_density(alpha=0.3) + xlab('Score') +
         geom_vline(xintercept = mean(plot_data$prob[plot_data$SFARI]), color = '#00C0C2', linetype = 'dashed') +
         geom_vline(xintercept = mean(plot_data$prob[!plot_data$SFARI]), color = '#FF7371', linetype = 'dashed') +
         theme_minimal() + ggtitle('Model score distribution by SFARI Label'))
```

- There seems to be a msall but consistent positive relation between the SFARI scores and the probability assigned by the model

```{r}
plot_data = test_set %>% mutate(ID=rownames(test_set)) %>% dplyr::select(ID, prob) %>%
            left_join(original_dataset %>% mutate(ID=rownames(original_dataset)), by='ID') %>%
            dplyr::select(ID, prob, gene.score) %>% apply_labels(gene.score='SFARI Gene score')

cro(plot_data$gene.score)

mean_vals = plot_data %>% group_by(gene.score) %>% summarise(mean_prob = mean(prob))

ggplotly(plot_data %>% ggplot(aes(gene.score, prob, fill=gene.score)) + geom_boxplot() + 
              scale_fill_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + 
              ggtitle('Distribution of probabilities by SFARI score') +
              xlab('SFARI score') + ylab('Probability') + theme_minimal())

rm(mean_vals)
```

Genes with highest scores in test set

<br>

- Considering the class imbalance in the test set (1:17), there are many more SFARI scores in here (1:3)

- 3 of the 5 SFARI genes with a score of 1 in the test set are in this top 50 list

- There aren't any SFARI genes with a score of 6

```{r}
test_set %>% dplyr::select(prob, SFARI) %>% mutate(ID = rownames(test_set)) %>% 
             arrange(desc(prob)) %>% top_n(50, wt=prob) %>%
             left_join(original_dataset %>% mutate(ID=rownames(original_dataset)), by='ID')  %>% 
             left_join(gene_names, by = c('ID'='ensembl_gene_id')) %>%
             dplyr::rename('GeneSymbol' = external_gene_id, 'Probability' = prob, 'ModuleDiagnosis_corr' = MTcor, 'GeneSignificance' = GS) %>%
             mutate(ModuleDiagnosis_corr = round(ModuleDiagnosis_corr,4), Probability = round(Probability,4), 
                    GeneSignificance = round(GeneSignificance,4)) %>%
             dplyr::select(GeneSymbol, GeneSignificance, ModuleDiagnosis_corr, Module, Probability, gene.score) %>%
             kable(caption = 'Genes with highest model probabilities from the test set')
```
<br>

---

<br><br>

### Negative samples distribution

<br>

Selecting the Negative samples in the test set
```{r}
negative_set = test_set %>% filter(!SFARI)

negative_set_table = negative_set %>% apply_labels(prob = 'Assigned Probability', 
                                                   pred = 'Label Prediction')

```

```{r}
cro(negative_set_table$pred)

cat(paste0('\n', sum(negative_set$pred), ' genes are predicted as ASD-related'))
```

```{r}
negative_set %>% ggplot(aes(prob)) + geom_density(color='#F8766D', fill='#F8766D', alpha=0.5) +
                 geom_vline(xintercept=0.5, color='#333333', linetype='dotted') + 
                 ggtitle('Probability distribution of the Negative samples in the test set') + 
                 theme_minimal()
```
<br>

#### Probability and Gene Significance

<br>

- There's a lot of noise, but the probability the model assigns to each gene seems to have a negative relation with the Gene Significance (under-expressed genes having on average the higher probabilities and over-expressed genes the lowest)

- The pattern is strongest in under-expressed genes

```{r, message=FALSE}
negative_set %>% ggplot(aes(prob, GS, color=MTcor)) + geom_point() + geom_smooth(method='loess', color='#666666') +
                 geom_hline(yintercept=0, color='gray', linetype='dashed') + xlab('Probability') +
                 scale_color_gradientn(colours=c('#F8766D','white','#00BFC4')) + 
                 ggtitle('Relation between Probability and Gene Significance') + theme_minimal()
```

<br>

#### Probability and Module-Diagnosis correlation

<br>

On average, the model seems to be assigning a probability inversely proportional to the Module-Diagnosis correlation of the module, with the highest positively correlated modules having the lowest average probability and the highest negatively correlated modules the highest average probability. But the difference isn't big

```{r probability_and_MTcor, fig.width=10, message=FALSE}
negative_set %>% ggplot(aes(MTcor, prob, color=GS)) + geom_point() + geom_smooth(method='loess', color='#666666') + 
                 geom_hline(yintercept=mean(negative_set$prob), color='gray', linetype='dashed') +
                 scale_color_gradientn(colours=c('#F8766D','#F8766D','white','#00BFC4','#00BFC4')) + 
                 xlab('Modules ordered by their correlation to ASD') + ylab('Model probability') +
                 theme_minimal()
```

Summarised version, plotting by module instead of by gene

The difference in the trend lines between this plot and the one above is that the one above takes all the points into consideration while this considers each module as an observation by itself, so the top one is strongly affected by big modules and the bottom one treats all modules the same

The model seems to give higher probabilities to genes belonging to modules with a small (absolute) correlation to Diagnosis (this is unexpected)

```{r, warning=FALSE, message=FALSE}
plot_data = negative_set %>% group_by(MTcor) %>% summarise(mean = mean(prob), sd = sd(prob), n = n()) %>%
            mutate(MTcor_sign = ifelse(MTcor>0, 'Positive', 'Negative')) %>% left_join(original_dataset, by='MTcor') %>%
            dplyr::select(Module, MTcor, MTcor_sign, mean, sd, n) %>% distinct()
colnames(plot_data)[1] = 'ID'

ggplotly(plot_data %>% ggplot(aes(MTcor, mean, size=n, color=MTcor_sign)) + geom_point(aes(id=ID)) + 
         geom_smooth(method='loess', color='gray', se=FALSE) + geom_smooth(method='lm', se=FALSE) + 
         xlab('Module-Diagnosis correlation') + ylab('Mean Probability by Model') + theme_minimal())
```
<br>

#### Probability and level of expression

<br>

There is a positive relation between level of expression and probability, the model seems to be capturing indirectly the level of expression of the genes to make the prediction, so it's introducing the same bias 

```{r, message=FALSE, warning=FALSE}
# Gandal dataset
load('./../Data/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame
```

```{r probability_and_meanExpr, warning=FALSE, message=FALSE}
mean_and_sd = data.frame(ID=rownames(datExpr), meanExpr=rowMeans(datExpr), sdExpr=apply(datExpr,1,sd))

plot_data = negative_set %>% left_join(mean_and_sd, by='ID') %>% 
            left_join(original_dataset %>% mutate(ID=rownames(original_dataset)) %>% 
                      dplyr::select(ID, Module), by='ID')
colnames(plot_data)[ncol(plot_data)] = 'Module'

plot_data %>% ggplot(aes(meanExpr, prob)) + geom_point(alpha=0.2, color='#0099cc') + 
              geom_smooth(method='loess', color='gray', alpha=0.3) + 
              geom_smooth(method='lm', color='#999999', se=FALSE, alpha=1) + 
              theme_minimal() + ggtitle('Mean expression vs model probability by gene')

rm(mean_and_sd)
```

```{r, message=FALSE}
plot_data2 = plot_data %>% group_by(Module) %>% summarise(meanExpr = mean(meanExpr), meanProb = mean(prob), n=n())

ggplotly(plot_data2 %>% ggplot(aes(meanExpr, meanProb, size=n)) + geom_point(color=plot_data2$Module) + 
         geom_smooth(method='loess', se=TRUE, color='gray', alpha=0.1, size=0.7) + 
         geom_smooth(method='lm', se=FALSE, color='gray') + theme_minimal() + theme(legend.position='none') + 
         ggtitle('Mean expression vs model probability by Module'))

rm(plot_data2)
```
<br>

#### Probability and lfc

<br>

There is a relation between probability and lfc, so it **IS*() capturing a bit of true information (because lfc and mean expression were negatively correlated and it still has a positive relation in the model)

- The relation is stronger in genes under-expressed in ASD

```{r probabilty_and_lfc, message=FALSE, fig.width=10}
plot_data = negative_set %>% left_join(DE_info %>% mutate(ID=rownames(DE_info)), by='ID')

plot_data %>% ggplot(aes(log2FoldChange, prob)) + geom_point(alpha=0.1, color='#0099cc') + 
              geom_smooth(method='loess', color='gray', alpha=0.3) + 
              theme_minimal() + ggtitle('lfc vs model probability by gene')
```

- The relation is stronger in Differentially Expressed genes

```{r probabilty_and_lfc_by_DE, message=FALSE, fig.width=10}
p1 = plot_data %>% filter(log2FoldChange<0) %>% mutate(DE = padj<0.05) %>% ggplot(aes(log2FoldChange, prob, color=DE)) + geom_point(alpha=0.1) + 
                   geom_smooth(method='loess', alpha=0.1) + xlab('') + ylab('Probability') + 
                   ylim(c(min(plot_data$prob), max(plot_data$prob))) + 
                   theme_minimal() + theme(legend.position = 'none', plot.margin=unit(c(1,-0.3,1,1), 'cm'))

p2 = plot_data %>% filter(log2FoldChange>=0) %>% mutate(DE = padj<0.05) %>% ggplot(aes(log2FoldChange, prob, color=DE)) + geom_point(alpha=0.1) + 
                   geom_smooth(method='loess', alpha=0.1) + xlab('') + ylab('Probability') + ylab('') +
                   scale_y_continuous(position = 'right', limits = c(min(plot_data$prob), max(plot_data$prob))) +
                   theme_minimal() + theme(plot.margin = unit(c(1,1,1,-0.3), 'cm'), axis.ticks.y = element_blank())

grid.arrange(p1, p2, nrow=1, top = 'LFC vs model probability by gene', bottom = 'LFC')

rm(p1, p2)
```
<br><br>

### Conclusion

<br>

The model is capturing the mean level of expression of the genes (indirectly through module memberhsip), which is a strong bias found in the SFARI scores, but it seems to be capturing a bit of true biological signal as well (based on the GS and the log fold change plots) 

<!-- ... -->

<!-- **New conclusion:** The model seems to be capturing some sort of confounding variable to make the predictions, it would seem that it's related to the mean expression or the standard deviation of the genes, but in `10_10_20_data_preprocessing_standardising_expr.RData` we standardised the dataset and it made no difference in the resulting clusterings, which means that there is some other behaviour related to the level of expression of a gene or its standard deviation that is biasing the classifier, but I don't know what it could be or how to fix it ... -->

---

#### Saving results

```{r save_results}
predictions = test_set

save(predictions, dataset, file='./../Data/Ridge_model_robust.RData')
```
<br><br>

---

#### Session info

```{r print_session_info}
sessionInfo()
```
<br><br>
