---
title: 'Dataset Exploratory Analysis'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, warning=FALSE, message=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-Models/FirstPUModel/RMarkdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(GGally)
library(WGCNA)
library(expss)
library(polycor)
library(knitr)

SFARI_colour_hue = function(r) {
  pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3','#808080','gray','#d9d9d9')[r]
}

clustering_selected = 'DynamicHybrid'

print(paste0('Using clustering ', clustering_selected))
```

Load preprocessed dataset (preprocessing code in 19_10_14_data_preprocessing.Rmd) and clustering (pipeline in 19_10_15_WGCNA.Rmd)
```{r load_dataset, message=FALSE, warning=FALSE}

# Gandal dataset
load('./../Data/Gandal/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame


# GO Neuronal annotations
GO_annotations = read.csv('./../../../PhD-InitialExperiments/FirstYearReview/Data/GO_annotations/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


# SFARI Genes
SFARI_genes = read_csv('./../Data/SFARI/SFARI_genes_08-29-2019_with_ensembl_IDs.csv')
SFARI_genes = SFARI_genes[!duplicated(SFARI_genes$ID) & !is.na(SFARI_genes$ID),]


# Update DE_info with SFARI and Neuronal information
genes_info = DE_info %>% mutate('ID'=rownames(.)) %>% left_join(SFARI_genes, by='ID') %>% 
  mutate(`gene-score`=ifelse(is.na(`gene-score`), 'None', `gene-score`)) %>%
  left_join(GO_neuronal, by='ID') %>% 
  mutate(Neuronal=ifelse(is.na(Neuronal), 0, Neuronal)) %>%
  mutate(gene.score=ifelse(`gene-score`=='None' & Neuronal==1, 'Neuronal', `gene-score`), 
         significant=padj<0.05 & !is.na(padj))


# Dataset created with DynamicTreeMerged algorithm
dataset = read.csv(paste0('./../Data/Gandal/dataset_', clustering_selected, '.csv'))
dataset$Module = dataset[,clustering_selected]


# GO DEA
# load('./../Data/Gandal/GO_DE_clusters.RData')

rm(DE_info, GO_annotations, clusterings)
```

```{r}
print(paste0('There are ', length(unique(dataset$Module)),' modules:'))

dataset$Module %>% table %>% sort(decreasing=TRUE) %>% print
```

---

## Module-Trait associations

```{r, fig.width=10, fig.height=10, warning=FALSE}
datTraits = datMeta %>% dplyr::select(Diagnosis_, Region, Sex, Age, PMI, RNAExtractionBatch) %>%
            rename('Diagnosis' = Diagnosis_, 'ExtractionBatch' = RNAExtractionBatch)

# Recalculate MEs with color labels
ME_object = datExpr %>% t %>% moduleEigengenes(colors = dataset$Module)
MEs = orderMEs(ME_object$eigengenes)

# Calculate correlation between eigengenes and the traits and their p-values
moduleTraitCor = MEs %>% apply(2, function(x) hetcor(x, datTraits, ML=TRUE)$correlations[1,-1]) %>% t
rownames(moduleTraitCor) = colnames(MEs)
colnames(moduleTraitCor) = colnames(datTraits)
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nrow(datExpr))

# In case there are any NAs
if(sum(!complete.cases(moduleTraitCor))>0){
  print(paste0(sum(is.na(moduleTraitCor)),' correlation(s) could not be calculated')) 
}
moduleTraitCor = moduleTraitCor[complete.cases(moduleTraitCor),]
moduleTraitPvalue = moduleTraitPvalue[complete.cases(moduleTraitCor),]

# Sort moduleTraitCor by Diagnosis
moduleTraitCor = moduleTraitCor[order(moduleTraitCor[,1], decreasing=TRUE),]
moduleTraitPvalue = moduleTraitPvalue[order(moduleTraitCor[,1], decreasing=TRUE),]

# Create text matrix for the Heatmap
textMatrix = paste0(signif(moduleTraitCor, 2), ' (', signif(moduleTraitPvalue, 1), ')')
dim(textMatrix) = dim(moduleTraitCor)


labeledHeatmap(Matrix = moduleTraitCor, xLabels = names(datTraits), yLabels =  gsub('ME','',rownames(moduleTraitCor)), 
               yColorWidth=0, colors = brewer.pal(11,'PiYG'), bg.lab.y = gsub('ME','',rownames(moduleTraitCor)),
               textMatrix = textMatrix, setStdMargins = FALSE, cex.text = 0.8, cex.lab.y = 0.75, zlim = c(-1,1),
               main = paste('Module-Trait relationships'))

rm(ME_object, datTraits, MEs, moduleTraitCor, moduleTraitPvalue, textMatrix)
```

Selecting the three modules with highest (absolute) correlation to Diagnosis: <span style="color:#D675FD">#D675FD</span>, <span style="color:#00BF78">#00BF78</span> and <span style="color:#13B700">#13B700</span>

```{r}
pca = datExpr %>% prcomp

plot_data = data.frame('ID'=rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>%
            left_join(dataset, by='ID') %>% dplyr::select(ID, PC1, PC2, Module, gene.score) %>%
            mutate(ImportantModules = ifelse(Module=='#D675FD', '#D675FD',
                                      ifelse(Module=='#00BF78', '#00BF78',
                                      ifelse(Module=='#13B700', '#13B700', 'Others')))) %>%
            mutate(color = ifelse(ImportantModules=='Others','gray',ImportantModules),
                   alpha = ifelse(ImportantModules=='Others', 0.1, 0.5))

table(plot_data$ImportantModules)

ggplotly(plot_data %>% ggplot(aes(PC1, PC2, color=ImportantModules)) + 
         geom_point(alpha=plot_data$alpha, color=plot_data$color) + theme_minimal() + 
           ggtitle('Modules with strongest relation to Diagnosis'))
```

<!-- GO enrichment analysis (analysis performed in 19_10_22_GO_enrichment_analysis.Rmd) -->
<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- for(mod in c('#C17FFF', '#00C094', '#C49A00')){ -->
<!--   if(nrow(EA_by_method[[clustering_selected]][[mod]])==0){ -->
<!--     print(paste0('There are no enriched terms for module ', mod)) -->
<!--   } else { -->
<!--     print(paste0('There are ', nrow(EA_by_method[[clustering_selected]][[mod]]),' enriched terms in module ', mod)) -->
<!--     kable(nrow(EA_by_method[[clustering_selected]][[mod]])) -->
<!--   } -->
<!-- } -->
<!-- ``` -->

---

## Gene Significance

Comparing this with the DEA plot, Gene significance close to 0 means the gene expression is not affected by Diagnosis and negative values means the gene is underexpressed in Autism samples. So **the important thing about gene significance is its absolute value**
```{r, warning=FALSE}
plot_data = data.frame('ID'=rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>%
            left_join(dataset, by='ID')

ggplotly(plot_data %>% ggplot(aes(PC1, PC2, color=GS)) + geom_point(alpha=0.5) + 
         scale_color_gradientn(colours=c('#F8766D','white','#00BFC4')) +
  theme_minimal() + ggtitle('Gene Significance'))
```

---

## Module Membership and Gene Significance relation

### Genes from the module <span style="color:#D675FD">#D675FD</span>

Module with the highest (absolute) correlation to Diagnosis (-0.93)
```{r}
plot_data = dataset %>% dplyr::select(MM.D675FD, GS, gene.score) %>% filter(dataset$Module=='#D675FD')

ggplotly(plot_data %>% ggplot(aes(MM.D675FD, GS, color=gene.score)) + geom_point(alpha=0.5) + 
         scale_color_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + 
         theme_minimal())
```


### Genes from the module <span style="color:#00BF78">#00BF78</span>

Module with the second highest (absolute) correlation to Diagnosis (0.91)
```{r}
plot_data = dataset %>% dplyr::select(MM.00BF78, GS, gene.score) %>% filter(dataset$Module=='#00BF78')

ggplotly(plot_data %>% ggplot(aes(MM.00BF78, GS, color=gene.score)) + geom_point(alpha=0.5) + 
         scale_color_manual(values=SFARI_colour_hue(r=c(1:6,8,7))) + 
         theme_minimal())
```


### Genes from the module <span style="color:#13B700">#13B700</span>

Module with the third highest (absolute) correlation to Diagnosis and the top one with positive correlation (0.91)
```{r}

plot_data = dataset %>% dplyr::select(MM.13B700, GS, gene.score) %>% filter(dataset$Module=='#13B700')


ggplotly(plot_data %>% ggplot(aes(MM.13B700, GS, color=gene.score)) + geom_point(alpha=0.5) + 
         scale_color_manual(values=SFARI_colour_hue(r=c(2:6,8,7))) + 
         theme_minimal())
```

---

## Identifying important genes

Selecting the modules with the highest correlation to Diagnosis, and, from them, the genes with the highest module membership-(absolute) gene significance

*Ordered by $\frac{MM+GS}{2}$

Only three SFARI Genes in the three lists ...
```{r select_important_genes, warning=FALSE, message=FALSE}
top_genes_1 = dataset %>% dplyr::select(ID, MM.D675FD, GS, gene.score) %>% filter(dataset$Module=='#D675FD') %>%
              rename('MM' = 'MM.D675FD') %>% mutate(importance = (MM+abs(GS))/2) %>% arrange(by=-importance) %>%
              top_n(10)

kable(top_genes_1, caption='Top 10 genes for module #D675FD')

top_genes_2 = dataset %>% dplyr::select(ID, MM.00BF78, GS, gene.score) %>% filter(dataset$Module=='#00BF78') %>%
              rename('MM' = 'MM.00BF78') %>% mutate(importance = (MM+abs(GS))/2) %>% arrange(by=-importance) %>%
              top_n(10)

kable(top_genes_2, caption='Top 10 genes for module #00BF78')

top_genes_3 = dataset %>% dplyr::select(ID, MM.13B700, GS, gene.score) %>% filter(dataset$Module=='#13B700') %>%
              rename('MM' = 'MM.13B700') %>% mutate(importance = (MM+abs(GS))/2) %>% arrange(by=-importance) %>%
              top_n(10)

kable(top_genes_3, caption='Top 10 genes for module #13B700')
```


```{r}
plot_data = data.frame('ID'=rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>%
            left_join(dataset, by='ID') %>% dplyr::select(ID, PC1, PC2, Module, gene.score) %>%
            mutate(color = ifelse(Module=='#D675FD', '#D675FD',
                           ifelse(Module=='#00BF78', '#00BF78',
                           ifelse(Module=='#13B700', '#13B700', 'gray')))) %>%
            mutate(alpha = ifelse(color %in% c('#D675FD', '#00BF78', '#13B700') & 
                                  ID %in% c(as.character(top_genes_1$ID), 
                                            as.character(top_genes_2$ID),
                                            as.character(top_genes_3$ID)), 1, 0.1))

plot_data %>% ggplot(aes(PC1, PC2)) + geom_point(alpha=plot_data$alpha, color=plot_data$color) + 
              theme_minimal() + ggtitle('Important genes identified through WGCNA')
```

---

## SFARI genes and Module-Diagnosis correlation

Modules with the strongest module-diagnosis correlation should have the highest percentage of SFARI Genes, but the plot shows this relation in modules with correlations close to 0, and the behaviour is the same for the other clusterings I have tried...

A good thing is that the gray cluster (which is actually all the genes that were left without a cluster) has a low correlation with ASD and also a low percentage of SFARI genes
```{r fig.width=10, warning=FALSE, message=FALSE}
plot_data = dataset %>% mutate('hasSFARIscore' = gene.score!='None') %>% 
            group_by(Module, MTcor, hasSFARIscore) %>% summarise(p=n()) %>% 
            left_join(dataset %>% group_by(Module) %>% summarise(n=n()), by='Module') %>% 
            mutate(p=round(p/n*100,2)) 

for(i in 1:nrow(plot_data)){
  this_row = plot_data[i,]
  if(this_row$hasSFARIscore==FALSE & this_row$p==100){
    new_row = this_row
    new_row$hasSFARIscore = TRUE
    new_row$p = 0
    plot_data = plot_data %>% rbind(new_row)
  }
}

plot_data = plot_data %>% filter(hasSFARIscore==TRUE)

ggplotly(plot_data %>% ggplot(aes(MTcor, p, size=n)) + geom_smooth(color='gray', se=FALSE) +
         geom_point(color=plot_data$Module, aes(id=Module)) +
         xlab('Module-Diagnosis correlation') + ylab('% of SFARI genes') +
         theme_minimal() + theme(legend.position = 'none'))

rm(i, this_row, new_row)
```
